---
title: "The epialleleR User's Guide"
author: "Oleksii Nikolaienko"
date: "`r format(Sys.time(), '%d %B, %Y')`"
abstract: |
  A comprehensive guide to using the epialleleR package for calling the
  hypermethylated epiallele frequencies from next-generation sequencing data
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{epialleleR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  width = 100
)
options(width=100)
devtools::load_all(".")   # delete later
```

*****

# Introduction

The cytosine DNA methylation is an important epigenetic mechanism for regulation
of gene expression. Abnormal methylation is linked to several diseases, being 
for example the most common molecular lesion in cancer cell[^1]. Multiple
studies suggest that alterations in DNA methylation, despite occuring at a low
mosaic level, may confer increased risk of cancer later in life[^2].

The cytosine methylation levels within relatively small regions of the human
genome are thought to be often concordant, resulting in a limited number of
distinct methylation patterns of short sequencing reads[^3]. Due to the
cell-to-cell variations in methylation, DNA purified from tissue samples
contains a mix of hyper- and hypomethylated alleles with varying ratios that
depend on a genomic region and a tissue type.

Unsurprisingly, when the frequencies of hypermethylated epialleles are low
(e.g. 1e-02 and lower) and cytosine methylation levels are averaged and reported
using conventional algorythms, the identification of such hypermethylated
epialleles becomes nearly impossible. In order to increase the sensitivity of
DNA methylation analysis we have developed *`epialleleR`* &#8212; an R package
for calling hypermethylated variant epiallele frequences (VEF).

*`epialleleR`* is a fast and scalable solution for analysis of data obtained by
next-generation sequencing of bisulfite-treated DNA samples. The minimum
requirement for the input is: Binary Alignment Map (BAM) file containing
sequencing reads. These reads can be obtained from either deep or ultra-deep
sequencing, using either narrowly targeted gene panel (amplicon sequencing),
larger methylation capture panels, or even whole-genome approaches.

## Current Features

 * in addition to conventional reporting of cytosine DNA methylation levels
 (beta values), *`epialleleR`* can call variant epiallele frequences (VEF) of
 hypermethylated alleles at the level of individual cytosines
 (*`generateCytosineReport`*) or supplied genomic regions
 (*`generateBedReport`*)
 * potential bimodality of methylation for genomic regions of interest can be
 assessed using *`generateBedEcdf`* method
 * the association of single-nucleotide variations with the epiallelic status
 can be tested using *`generateVcfReport`*

*****

# Sample data

*`epialleleR`* package includes sample data, which was obtained using targeted
sequencing. The description of assays and files is given below. All the
genomic coordinates for external data files are according to GRCh38 reference
assembly.

### Amplicon-based methylation NGS data

The samples of Human HCT116 DKO Non-Methylated (Zymo Research, cat # D5014-1),
or Human HCT116 DKO Methylated (Zymo Research, cat # D5014-2) DNA[^4], or their
mix were bisulfite-converted, and BRCA1 gene promoter was amplified using four
pairs of primers. Amplicons were mixed, indexed and sequenced at Illumina MiSeq
system. The related files are:

| Name | Type | Description |
| --- | --- | --- |
| amplicon000meth.bam | BAM | a subset of reads for non-methylated DNA sample |
| amplicon010meth.bam | BAM | a subset of reads for a 1:9 mix of methylated and non-methylated DNA samples |
| amplicon100meth.bam | BAM | a&nbsp;subset&nbsp;of&nbsp;reads&nbsp;for&nbsp;fully&nbsp;methylated&nbsp;DNA&nbsp;sample |
| amplicon.bed | BED | genomic coordinates of four amplicons covering promoter area of BRCA1 gene |
| amplicon.vcf.gz | VCF | a relevant subset of sequence variations |
| amplicon.vcf.gz.tbi | tabix | tabix file for the amplicon.vcf.gz |

### Capture-based methylation NGS data

The tumour DNA was bisulfite-converted, fragmented and hybridyzed with
custom-made probes covering promoter regions of 283 tumour supressor genes (as
described in [^5]). Libraries were sequenced using Illumina MiSeq system. The
related files are:

| Name | Type | Description |
| --- | --- | --- |
| capture.bam | BAM | a subset of reads |
| capture.bed | BED | genomic&nbsp;coordinates&nbsp;of&nbsp;capture&nbsp;target&nbsp;regions |
| capture.vcf.gz | VCF | a relevant subset of sequence variations |
| capture.vcf.gz.tbi | tabix | tabix file for the amplicon.vcf.gz |

*****

# Typical workflow

As mentioned earlier, *`epialleleR`* uses data stored in Binary Alignment Map
(BAM) files as its input. It is absolutely necessary that records in a BAM file
contain XM tag with the methylation call string &#8212; such files are produced
by virtually any software tool for mapping and alignment of bisulfite
sequencing reads (such as Bismark, BSMAP or Illumina DRAGEN Bio-IT Platform).

Please use function help files for full description of available parameters, as
well as explanation of function's logic and output values.

## Reading the data

All of the *`epialleleR`* methods can load BAM data using the file path.
However if a file is very large and several reports need to be prepared, it is
advised to use *`preprocessBam`* convenience function as shown below. This
function is also used internally when BAM file location string is supplied as
an input for other *`epialleleR`* methods.

*`preprocessBam`* automatically determines if BAM is derived from single-end or
paired-end sequencing. When the latter is the case, paired reads are merged
so that the overlapping fragments of second read are clipped (because quality
of the second read is usually lower than of the first). These **merged reads**
are then processed as a **single entity** in all *`epialleleR`* methods.

```{r}
library(epialleleR)

capture.bam <- system.file("extdata", "capture.bam", package="epialleleR")
bam.data    <- preprocessBam(capture.bam)
```

## Making cytosine reports

*`epialleleR`* can generate conventional cytosine reports in a format, which is
similar to the genome-wide cytosine report produced by coverage2cytosine
Bismark module[^6].

Please note that *`generateCytosineReport`* produces thresholded (VEF) report
by default: **methylated** cytosines from reads that do **not** pass the
threshold (**hypo**methylated reads) are counted as being **un**methylated. In
order to make conventional cytosine report, use *`threshold.reads=FALSE`*.

```{r}
# data.table::data.table object for
# CpG VEF report
cg.vef.report <- generateCytosineReport(bam.data)
head(cg.vef.report[order(meth+unmeth, decreasing=TRUE)])

# CpG cytosine report
cg.report <- generateCytosineReport(bam.data, threshold.reads=FALSE)
head(cg.report[order(meth+unmeth, decreasing=TRUE)])

# CX cytosine report
cx.report <- generateCytosineReport(bam.data, threshold.reads=FALSE,
                                    report.context="CX")
head(cx.report[order(meth+unmeth, decreasing=TRUE)])
```

## Making VEF reports for the set of genomic regions

*`epialleleR`* allows to make reports not only for individual cytosine bases,
but also for a set of genomic regions. It is especially useful when the targeted
methylation sequencing was used to produce reads (such as amplicon sequencing or
by capture using, e.g., Agilent SureSelect Target Enrichment Probes).

The amplicon sequencing principally differs from capture-based assays in that
the coordinates of reads are known. Therefore reads can be assigned to amplicons
by their exact positions, while to the capture targets &#8212; by the overlap.
For this, *`epialleleR`* provides generic *`generateBedReport`* function as well
as two of its aliases, *`generateAmpliconReport`* (for amplicon-based NGS) and
*`generateCaptureReport`* (for capture-based NGS).

```{r}
# report for amplicon-based data
# matching is done by exact start or end positions plus/minus tolerance
amplicon.report <- generateAmpliconReport(
  bam=system.file("extdata", "amplicon010meth.bam", package="epialleleR"),
  bed=system.file("extdata", "amplicon.bed", package="epialleleR")
)
amplicon.report

# report for capture-based data
# matching is done by overlap
capture.report <- generateCaptureReport(
  bam=system.file("extdata", "capture.bam", package="epialleleR"),
  bed=system.file("extdata", "capture.bed", package="epialleleR")
)
head(capture.report)

# generateBedReport is a generic function for BED-guided reports
bed.report <- generateBedReport(
  bam=system.file("extdata", "capture.bam", package="epialleleR"),
  bed=system.file("extdata", "capture.bed", package="epialleleR"),
  bed.type="capture"
)
identical(capture.report, bed.report)
```

## Plotting the distribution of per-read beta values

As stated in the introduction, human genomic DNA regions often show bimodal
methylation patterns. *`epialleleR`* allows to visualize this information by
plotting empirical cumulative distribution functions (eCDFs) for within- and
out-of-context beta values.

The code below produces plots for the sequencing reads of control DNA
samples. Note that within-the-context eCDF(0.5) values are very close to the
expected 1-VEF values for the corresponding control DNA samples: 

 * non-methylated DNA &#8212; expected VEF = 0, observed 1-eCDF(0.5) &#8776; 0
 * 1:9 mix of methylated and non-methylated DNA &#8212; expected VEF = 0.1, observed 1-eCDF(0.5) &#8776; 0.1
 * and fully methylated DNA &#8212; expected VEF = 1, observed 1-eCDF(0.5) &#8776; 1

```{r, fig.width=10, fig.height=6, out.width="100%", out.height="100%"}
# First, let's visualise eCDFs for within- and out-of-context beta values
# for all four amplicons and unmatched reads. Note that within-the-context eCDF
# of 0.5 is very close to the expected 1-VEF value (0.1) for all amplicons
# produced from this 1:9 mix of methylated and non-methylated control DNA

# let's compute eCDF
amplicon.ecdfs <- generateBedEcdf(
  bam=system.file("extdata", "amplicon010meth.bam", package="epialleleR"),
  bed=system.file("extdata", "amplicon.bed", package="epialleleR"),
  bed.rows=NULL
)

# there are 5 items in amplicon.ecdfs, let's plot all of them
par(mfrow=c(1,length(amplicon.ecdfs)))

# cycle through items
for (x in 1:length(amplicon.ecdfs)) {
  # four of them have names corresponding to genomic regions of amplicon.bed
  # fifth - NA for all the reads that don't match to any of those regions
  main <- if (is.na(names(amplicon.ecdfs[x]))) "unmatched"
          else names(amplicon.ecdfs[x])
  
  # plotting eCDF for within-the-context per-read beta values (in red)
  plot(amplicon.ecdfs[[x]]$context, col="red", verticals=TRUE, do.points=FALSE,
       xlim=c(0,1), xlab="per-read beta value", ylab="cumulative density", main=main)
  
  # adding eCDF for out-of-context per-read beta values (in blue)
  plot(amplicon.ecdfs[[x]]$out.of.context, add=TRUE, col="blue",
       verticals=TRUE, do.points=FALSE)
}


# Second, let's compare eCDFs for within-the-context beta values for only one
# amplicon but all three sequenced samples: pure non-methylated DNA, 1:9 mix of
# methylated and non-methylated DNA, and fully methylated DNA

# our files
bam.files <- c("amplicon000meth.bam", "amplicon010meth.bam", "amplicon100meth.bam")

# let's plot all of them
par(mfrow=c(1,length(bam.files)))

# cycle through items
for (f in bam.files) {
  # let's compute eCDF
  amplicon.ecdfs <- generateBedEcdf(
    bam=system.file("extdata", f, package="epialleleR"),
    bed=system.file("extdata", "amplicon.bed", package="epialleleR"),
    # only the second amplicon
    bed.rows=2, verbose=FALSE
  )
  
  # plotting eCDF for within-the-context per-read beta values (in red)
  plot(amplicon.ecdfs[[1]]$context, col="red", verticals=TRUE, do.points=FALSE,
       xlim=c(0,1), xlab="per-read beta value", ylab="cumulative density", main=f)
  
   # adding eCDF for out-of-context per-read beta values (in blue)
  plot(amplicon.ecdfs[[1]]$out.of.context, add=TRUE, col="blue",
       verticals=TRUE, do.points=FALSE)
}
```

## Exploring sequence variants in epialleles

It is known that sequence variants can affect methylation status of a DNA[^7].
*`generateVcfReport`* function calculates frequencies of single nucleotide
variants (SNVs) within epialleles and tests for the association between SNV and
epiallelic status using Fisher's exact test. Base counts and the test's p-values
are included in the returned value.

In addition to BAM file location string or preprocessed BAM object, function
requires a location string for the Variant Call Format (VCF) file or the VCF
object that was obtained using *`VariantAnnotation::readVcf`* function. As VCF
files can be extremely large, it is strongly advised to prefilter the VCF object
by the relevant set of genomic regions, or specify such relevant set of regions
as a *`bed`* parameter when *`vcf`* points to a VCF file location.

Please note, that the output report is currently limited to SNVs only.

```{r}
# VCF report
vcf.report <- generateVcfReport(
  bam=system.file("extdata", "amplicon010meth.bam", package="epialleleR"),
  bed=system.file("extdata", "amplicon.bed", package="epialleleR"),
  vcf=system.file("extdata", "amplicon.vcf.gz", package="epialleleR"),
  # when VCF seqlevels are different from BED and BAM it is possible
  # to convert them internally
  vcf.style="NCBI"
)

# NA values are shown for the C->T variants on the "+" and G->A on the "-"
# strands, because bisulfite conversion makes their counting impossible
head(vcf.report)

# let's sort the report by increasing Fisher's exact test's p-values.
# the p-values are given separately for reads that map to the "+"
head(vcf.report[order(`FEp-`, na.last=TRUE)])

# and to the "-" strand
head(vcf.report[order(`FEp+`, na.last=TRUE)])
```

*****

# Other information

## Citing the *`epialleleR`* package

The experimental data analysed using the package has not been published yet.
The citation information will be updated in the nearest future.


## Session Info

```{r session}
sessionInfo()
```

*****

# References

[^1]: https://doi.org/10.1146/annurev.pharmtox.45.120403.095832
[^2]: https://doi.org/10.1101/2020.12.01.403501
[^3]: https://doi.org/10.1093/bib/bbx077
[^4]: https://www.zymoresearch.com/products/human-methylated-non-methylated-dna-set-dna-w-primers
[^5]: https://dx.doi.org/10.1186%2Fs13148-020-00920-7
[^6]: https://www.bioinformatics.babraham.ac.uk/projects/bismark/
[^7]: https://doi.org/10.1038/modpathol.2009.130

