---
title: "The epialleleR output values <img style=\"float: right;\" src=\"epialleleR_logo.svg\"> "
date: "`r format(Sys.time(), '%d %B, %Y')`"
abstract: |
  A comparison and visualisation of epialleleR output values for various input
  files
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{values}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  width = 100
)
options(width=100)
```

# Introduction

The best possible explanation on VEF and lMHL values is given in help files for
*`generateCytosineReport`* and *`generateMhlReport`* methods, respectively.
Here we try to show some simplified and real situations,
i.e., different methylation patterns that may exist, and provide a visual
summary of *`epialleleR`* output.

The readers are welcome to try their own real and simulated data. If it might
be of interest to others, please create an issue and these examples might get
included in this vignette.

NB: the `plotMetrics` function used below is a piece of spaghetti code, hence
hidden. If you still want to use it or see what it does - browse a
[source code](https://github.com/BBCG/epialleleR/blob/devel/vignettes/values.Rmd)
of this vignette online.


```{r, echo=FALSE, include=FALSE}
require("data.table", quietly=TRUE)
require("GenomicRanges", quietly=TRUE)
require("ggplot2", quietly=TRUE)
require("epialleleR", quietly=TRUE)
  
plotMetrics <- function (bam.file, range, min.n=0, title="epialleles") {
  bam <- preprocessBam(bam.file=bam.file, verbose=FALSE)
  cg.beta <- generateCytosineReport(bam, threshold.reads=FALSE, verbose=FALSE)
  cg.vef  <- generateCytosineReport(bam, threshold.reads=TRUE, verbose=FALSE)
  cg.mhl  <- generateMhlReport(bam, max.haplotype.window=20, verbose=FALSE)
  range.strand <- as.character(strand(range))
  if (range.strand=="*") range.strand <- c("+", "-")
  metrics <- cbind(
    cg.beta[rname==as.character(seqnames(range)) & pos>=start(range) & pos<=end(range) & strand %in% range.strand,
            .(pos=factor(pos), beta=meth/(meth+unmeth))],
    cg.vef[rname==as.character(seqnames(range)) & pos>=start(range) & pos<=end(range)  & strand %in% range.strand,
           .(VEF=meth/(meth+unmeth))],
    cg.mhl[rname==as.character(seqnames(range)) & pos>=start(range) & pos<=end(range)  & strand %in% range.strand,
           .(lMHL=lmhl)]
  )

  metrics.melt <- melt.data.table(metrics, id.vars="pos")
  metrics.melt[value<0.00001, value:=0]
  metrics.melt[, pos:=as.numeric(as.character(pos))]
  
  grid::grid.newpage()
  patterns <- extractPatterns(bam, range, verbose=FALSE)[strand %in% range.strand]
  ctx.size <- 3 - findInterval(ncol(patterns), c(20, 50))
  epi.grob <- plotPatterns(patterns, marginal="count", npatterns.per.bin=Inf, context.size=ctx.size,
                           plot=FALSE, verbose=FALSE, title=title, subtitle=NULL)
  
  values <- ggplot(metrics.melt,
                   aes(x=pos, y=value, color=variable, group=variable)) +
    geom_line(alpha=0.5, linewidth=1) +
    scale_color_brewer(palette="Set1") +
    scale_x_continuous(name=NULL, breaks=c()) +
    scale_y_continuous(position="right", name=NULL, trans="log10", limits=c(0.00001,1), breaks=10**-c(0:5)) +
    theme_light() +
    theme(legend.position="right")
  
  nul.grob <- ggplotGrob(ggplot()+theme_void())
  nul.grob$widths[[7]] <- grid::unit(0.25, "null")
  val.grob <- ggplotGrob(values)
  full.plot <- rbind(epi.grob, cbind(nul.grob, val.grob))
  grid::grid.draw(full.plot);
}

```


```{r, fig.width=10, fig.height=8, out.width="100%", out.height="100%", warning=FALSE}
out.bam <- tempfile(pattern="simulated", fileext=".bam")
set.seed(1)

# no epimutations
simulateBam(
  output.bam.file=out.bam,
  XM=c(
    sapply(
      lapply(1:1000, function (x) sample(c("Z",rep("z", 9)), 10)),
      paste, collapse=""
    )
  ),
  XG="CT"
)
plotMetrics(out.bam, as("chrS:1-10", "GRanges"), 0, title="no epimutations")

# one complete epimutation
simulateBam(
  output.bam.file=out.bam,
  XM=c(
    paste(rep("Z", 10), collapse=""),
    sapply(
      lapply(1:999, function (x) sample(c("Z",rep("z", 9)), 10)),
      paste, collapse=""
    )
  ),
  XG="CT"
)
plotMetrics(out.bam, as("chrS:1-10", "GRanges"), title="one complete epimutation")

# one partial epimutation
simulateBam(
  output.bam.file=out.bam,
  XM=c(
    paste(c(rep("Z", 4), "z", "z", rep("Z", 4)), collapse=""),
    sapply(
      lapply(1:999, function (x) sample(c("Z",rep("z", 9)), 10)),
      paste, collapse=""
    )
  ),
  XG="CT"
)
plotMetrics(out.bam, as("chrS:1-10", "GRanges"), title="one partial epimutation")

# another partial epimutation
simulateBam(
  output.bam.file=out.bam,
  XM=c(
    "zZZZZZZZzz",
    sapply(
      lapply(1:999, function (x) sample(c("Z",rep("z", 9)), 10)),
      paste, collapse=""
    )
  ),
  XG="CT"
)
plotMetrics(out.bam, as("chrS:1-10", "GRanges"), title="another partial epimutation")

# several partial epimutations
simulateBam(
  output.bam.file=out.bam,
  XM=c(
    sapply(
      lapply(1:10, function (x) c(rep("Z", 6), rep("z", 4))),
      paste, collapse=""
    ),
    sapply(
      lapply(1:999, function (x) sample(c("Z",rep("z", 9)), 10)),
      paste, collapse=""
    )
  ),
  XG="CT"
)
plotMetrics(out.bam, as("chrS:1-10", "GRanges"), title="several partial epimutations")

# several short partial epimutations
simulateBam(
  output.bam.file=out.bam,
  XM=c(
    sapply(
      lapply(1:10, function (x) c(rep("Z", 4), rep("z", 6))),
      paste, collapse=""
    ),
    sapply(
      lapply(1:999, function (x) sample(c("Z",rep("z", 9)), 10)),
      paste, collapse=""
    )
  ),
  XG="CT"
)
plotMetrics(out.bam, as("chrS:1-10", "GRanges"), title="several short partial epimutations")

# several overlapping partial epimutations
simulateBam(
  output.bam.file=out.bam,
  pos=1:10,
  XM=c(
    "ZZZZZZZZZZ", "ZZZZZZZZZz", "ZZZZZZZZzz", "ZZZZZZZzzz", "ZZZZZZzzzz",
    sapply(
      lapply(1:15, function (x) sample(c("Z",rep("z", 9)), 10)),
      paste, collapse=""
    )
  ),
  XG="CT"
)
plotMetrics(out.bam, as("chrS:1-20", "GRanges"), title="several overlapping partial epimutations")

# amplicon 0%
plotMetrics(
  system.file("extdata", "amplicon000meth.bam", package="epialleleR"),
  as("chr17:43124861-43126026", "GRanges"), title="amplicon, 0%"
)

# amplicon 10%
plotMetrics(
  system.file("extdata", "amplicon010meth.bam", package="epialleleR"),
  as("chr17:43124861-43126026", "GRanges"), title="amplicon, 10%"
)

# sample capture, BMP7
plotMetrics(
  system.file("extdata", "capture.bam", package="epialleleR"),
  as("chr20:57266125-57268185:+", "GRanges"), title="sample capture, BMP7, + strand"
)

# sample capture, BMP7
plotMetrics(
  system.file("extdata", "capture.bam", package="epialleleR"),
  as("chr20:57266125-57268185:-", "GRanges"), title="sample capture, BMP7, - strand"
)

# sample capture, RAD51C
plotMetrics(
  system.file("extdata", "capture.bam", package="epialleleR"),
  as("chr17:58691673-58693108:+", "GRanges"), title="sample capture, RAD51C, + strand"
)

# sample capture, RAD51C
plotMetrics(
  system.file("extdata", "capture.bam", package="epialleleR"),
  as("chr17:58691673-58693108:-", "GRanges"), title="sample capture, RAD51C, - strand"
)

# long-read sequencing, low methylation
getXM <- function (p) {sample(x=c("z", "Z"), size=1, prob=c(p, 1-p))}
probs <- (sin(seq(-2*pi, +1*pi, by = pi/25))+2)/3
simulateBam(
  output.bam.file=out.bam,
  pos=1:10,
  XM=sapply(1:10, function (i) {paste(sapply(probs, getXM), collapse="")}),
  XG="CT"
)
plotMetrics(out.bam, as("chrS:1-1000", "GRanges"), title="long-read sequencing, low methylation")

# long-read sequencing, high methylation
simulateBam(
  output.bam.file=out.bam,
  pos=1:10,
  XM=sapply(1:10, function (i) {paste(sapply(1-probs, getXM), collapse="")}),
  XG="CT"
)
plotMetrics(out.bam, as("chrS:1-1000", "GRanges"), title="long-read sequencing, high methylation")

```


## Session Info

```{r session}
sessionInfo()
```

